### 2016-02-27 paper/icfp16-submission/semantics.tex
trascribing the extended semantics

\subsection{Syntax}
\label{sec:syntax}
\input{syntax.tex}
%
Figure~\ref{fig:syntax} describes the syntax of \lang, a simple lambda
calculus with integers and booleans.
%
As we are specifically interested in programs that \emph{do} go wrong,
we include an explicit \stuck\ state in our syntax.

\paragraph{Holes}
\label{sec:holes}
%
Recall that a key challenge in our setting is to find witnesses
that are meaningful and do not arise from choosing values from
irrelevant types.
%
We solve this problem be equipping our term language with a notion
of a \emph{hole}, written \ehole{}, which represents \emph{unconstrained}
values that may replaced with values from \emph{any} type.
%
Dually, holes may also appear in types, where they may be thought
of as type variables that we will not generalize over.
%

@@

\subsection{Semantics}
\label{sec:semantics}
\input{operational.tex}
%

@@

\paragraph{Narrowing Types} The procedure $\force{e}{t}{\vsu}$ takes as
input an term \texttt{e} and refines it to have the type \texttt{t} by
yielding a pair of either the same expression and value
substitution, or yields the stuck term if no such refinement is
possible. In the case where \texttt{e} is a hole, it first checks in the
given $\vsu$ to see if the hole has already been instantiated and, if
so, returns the existing instantiation:
%
$$
\begin{array}{lcl}
%% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
\forcesym            & ::     & (e, t, \vsu) \rightarrow \pair{e}{\vsu} \\
\force{v}{\ehole{}}{\vsu}  & \defeq & \pair{v}{\vsu} \\
\force{\ehole{i}}{t}{\vsu} & \defeq & \hspace{-1ex}
\begin{cases}
  \pair{\vsu(i)}{\vsu}        & \mbox{if}\ i \in dom(\vsu) \\
  \pair{v}{\ehole{i} \mapsto v ; \vsu}   & v = \gen{t}\\
\end{cases} \\
 % \begin{array}{l}
    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \pair{\vsu(i)}{\vsu}\ \mathtt{else} \\
         % \elet{v}{\gen{t}}{\pair{v}{\ehole{i} \mapsto v}}
  % \end{array} \\
\force{n}{\tint}{\vsu}     & \defeq & \pair{n}{\vsu} \\
\force{v}{\tint}{\vsu}     & \defeq & \pair{\stuck}{\vsu} \\
\force{b}{\tbool}{\vsu}    & \defeq & \pair{b}{\vsu} \\
\force{v}{\tbool}{\vsu}    & \defeq & \pair{\stuck}{\vsu} \\
\force{\efun{x}{e}}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\efun{x}{e}}{\vsu} \\
\force{v}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\stuck}{\vsu} \\
\end{array}
$$
%
While a hole may map to a value that \emph{contains} another hole, \eg a
lambda, it may not map \emph{directly} to another hole, so in the second
case above we do not need to \forcesym\ the result of the substitution,
the sub-hole will be narrowed when the context demands it.

\paragraph{Generating Values} The (non-deterministic) procedure $\gen{t}$
takes as input a type \texttt{t} and returns a value of that type. For base
types, the procedure returns an arbitrary value. For functions, it returns
a lambda with a \emph{new} hole denoting the return value, and unconstrained
types (denoted by $\thole{}$) yield fresh, unconstrained values
(denoted by $\ehole{i}$).

$$
\begin{array}{lcll}
% \gensym       & ::      & t \rightarrow v \\
\gen{\tint}   & \defeq  & n, & \quad \text{non-deterministic} \\
\gen{\tbool}  & \defeq  & b, & \quad \text{non-deterministic} \\
\gen{\tfun{t_1}{t_2}}   & \defeq & \efun{x}{\ehole{i}}, & \quad \text{$i$ is fresh} \\
\gen{\thole{}} & \defeq & \ehole{i}, & \quad \text{$i$ is fresh} \\
\end{array}
$$

\paragraph{Steps and Traces}
%
% WRW notes that Figure 4 does not seem to handle recursion (it's not clear
% how the let rule would work for something "let rec"-y, and there's not
% function call rule). I only mention this because I can imagine a reviewer
% wondering about your ability to generate good witnesses for function
% types. This could likely be addressed in text, by a forward reference to
% Section 3.4 where higher-order functions are handled, without changing
% any of the formalisms at the last minute.
Figure~\ref{fig:operational} describes the small-step contextual
reduction semantics for \lang.
%
A \emph{closed expression} is a pair $\pair{e}{\vsu}$ of an expression $e$
and a substitution $\vsu$.
%
We write $\step{e}{\vsu}{e'}{\vsu'}$ if the closed expression $\pair{e}{\vsu}$
transitions in a \emph{small step} to $\pair{e'}{\vsu'}$.
%
A (finite) \emph{trace} $\trace$ is a sequence of closed expressions
$\pair{e_0}{\vsu_0}, \ldots, \pair{e_n}{\vsu_n}$ such that
$\forall 0 \leq i < n$, we have \step{e_i}{\vsu_i}{e_{i+1}}{\vsu_{i+1}}.
%
We write \steptr{\trace}{e}{\vsu}{e'}{\vsu'} if $\trace$ is a trace of the form
$\pair{e}{\vsu},\ldots,\pair{e'}{\vsu'}$.
%
We write \steps{e}{\vsu}{e'}{\vsu'} if \steptr{\trace}{e}{\vsu}{e'}{\vsu'}
for some trace $\trace$.

\paragraph{Primitive Reductions}
Primitive reduction steps --- addition, if-elimination, and
function application --- use \forcesym to ensure that values have the
appropriate type (and that holes are instantiated) before continuing the
computation. Importantly, beta-reduction \emph{does not} type-check its
argument, it only ensures that $v_1$ is a function.

@@

\paragraph{Witness Generality}
We now show that our evaluation relation instantiates holes in a
\emph{general} manner. That is, given a function $f$, if we have
$\steps{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su}$, then for every
possible input type $t$, we can find a value $\hastype{v}{t}$
such that $\eapp{f}{v}$ goes wrong.

\begin{thm}{\textbf{[Witness Generality]}}
\label{thm:soundness}
  For any function $f$, if \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su},
  then for every inhabitable type\footnote{We exclude builtin functions that subvert the type system, \eg \texttt{Obj.magic}, and thus consider the type $\forall a b. \tfun{a}{b}$ to be uninhabitable.}  $t$ there exists $\hastype{v}{t}$ such that
  $\steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}$.
\end{thm}

@@

\paragraph{The Type of an Expression} The (dynamic) type of an
expression $e$ is defined as a function $\typeof{e}$ defined as:
  \[
  \begin{array}{lcll}
    \typeof{n}   & \defeq & \tint & \\
    \typeof{b}   & \defeq & \tbool & \\
    \typeof{\efun{x}{e}} & \defeq & \tfun{\thole{i}}{\typeof{e}}, & \quad \text{$i$ is fresh} \\
    \typeof{e} & \defeq & \thole{i}, & \quad \text{$i$ is fresh} \\
  \end{array}
  \]
%
Note that we approximate the type of expressions rather than
performing full type inference, for example
$\typeof{\efun{x}{\eplus{x}{1}}} = \tfun{\thole{}}{\thole{}}$,
instead of the usual $\tfun{\tint}{\tint}$.

\paragraph{Dynamic Type Inference}
We can think of the evaluation of \eapp{f}{\ehole{}} as (dynamically)
inferring a \emph{partial type} --- a type that may contain holes ---
for $f$.
%
We can extract this type from evaluation traces.
%
Formally, we say that if \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{e}{\su},
then the \emph{partial type} of $f$ upto $\trace$, written \ptype{\trace}{f},
is \tfun{\typeof{\subst{\su}{\ehole{}}}}{\typeof{e}}.
%
%We will omit the subscript when we wish to refer to the final partial
%type, \ie\ at the step where the expression has been reduced to a value
%(or stuck.)

\paragraph{Type Compatibility}
Two types are compatible if one of them can be obtained by
suitably filling the holes of the other (\ie by suitably ``instantiating''
the type variables appearing in the other). Formally, a type $s$ is
\emph{compatible} with a type $t$, written \tcompat{s}{t}, if
$\exists \tsu.\ t = \subst{\tsu}{s} \lor s = \subst{\tsu}{t}$.

\paragraph{Preservation}
Given these two definitions, we show that each evaluation step
refines the partial type of $f$, \ie preserves type compatibility.
%
\begin{lem}
\label{lem:refine-partial}
If $\trace \defeq \pair{\eapp{f}{\ehole{}}}{\emptysu},\ldots$ and
$\trace' \defeq \trace, \pair{e'}{\vsu'}$ (\ie $\trace'$ is a single-step
extension of $\trace$)
%
%The partial type of $f$ upto $\trace$ is compatible
%with the partial type upto $\trace'$, \ie\
%
then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
\end{lem}
\begin{proof}
  By induction on the length of $\trace$ and case analysis on the evaluation rules.
  %
  Note that all rules preserve partial types with the exception of when
  \forcesym\ is called on a hole, in which case we instantiate the hole with
  a concrete value.
  %
  But $\typeof{\ehole{}} = \thole{}$, which is compatible with any type.
\end{proof}

\paragraph{Narrowing}
%
Furthermore, only a call to \forcesym can change the partial type of $f$.
%
\begin{lem}
\label{lem:force-inst}
If
$\trace \defeq \pair{\eapp{f}{\ehole{}}}{\emptysu}, \ldots,\pair{e}{\vsu}$
and
$\trace' \defeq \trace, \step{e}{\vsu}{e'}{\vsu'}$
and
$\ptype{\trace}{f} \neq \ptype{\trace'}{f}$,
then the final step $\step{e}{\vsu}{e'}{\vsu'}$ invokes \forcesym.
\end{lem}

@@

\paragraph{Incompatible Values Are Wrong}
%
\emph{Any} value that is \emph{incompatible} with
the partial type upto trace $\trace$ will cause $f$ to get stuck
in \emph{at most} $k$ steps where $k$ is the length of $\trace$.
%
\begin{lem}
\label{lem:k-stuck}
  If \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{e}{\vsu} and
     \hastype{v}{t} and
     \tincompat{t}{\ptype{\trace}{f}},
  then
     \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
\end{lem}
\begin{proof}
  By induction on $k$, the length of $\trace$.
  Let $\tfun{s_{\trace}}{\_} = \ptype{\trace}{f}$ and
  $\trace' = \pair{\eapp{f}{\ehole{}}}{\emptysu},...,\pair{e'}{\vsu'}$ 
  such that $\step{e'}{\vsu'}{e}{\vsu}$.
  Suppose \hastype{v}{\tincompat{t}{s_{\trace}}}, we
  will show that \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su'}
  in at most $k$ steps:
%
  %\begin{description}
  % \item
  [Case \tincompat{t}{s_{\trace'}}:]
    The inductive hypothesis applies.
  %\item
  [Case $\tcompat{t}{s_{\trace'}}$ but $\tincompat{t}{s_{\trace}}$:]
    By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
    $k$.
    %
    Lemma~\ref{lem:refine-partial} tells us
    $\tcompat{s_{\trace'}}{s_{\trace}}$, which means we must have
    specifically narrowed $s_{\trace'}$ to a type incompatible with $t$.
    %
    A case analysis of the rules shows that \forcesym cannot have
    succeeded.
  %\end{description}
\end{proof}

@@

\paragraph{Modeling Semantics}
%: \texttt{eval} and \texttt{isStuck}}
%
We resolve the non-determinism in the operational semantics
(\S~\ref{sec:semantics}) via the procedure:
%
\begin{mcode}
  eval :: ($e$, $\su$) -> [($v$, $\su$)]
\end{mcode}
%
Due to the non-determinism introduced by \gensym, a call \hbox{|eval($e$, $\vsu$)|}
returns the \emph{list} of possible results of the form $\pair{e'}{\vsu'}$ such
that $\steps{e}{\vsu}{e'}{\vsu'}$.
%
We determine whether a given value is stuck by using the predicate:
%
\begin{mcode}
  isStuck :: $v$ -> Bool
\end{mcode}

@@

close :: [($x$, $e$)] -> [$v$] -> ($e$, $\vsu$)
close bs vs = (e, $\emptysu$)
  where
    e       = mkLets bs (witness bs vs)

@@

\begin{cor}{\textbf{[Witness Generation]}}
\label{thm:generation}
  If @genWitness k [(f1,e1),...,(fn,en)]@ is non-empty,
  then for every type $t$ there exists $\hastype{v}{t}$ such that
  $\steps{\eapp{f_n}{v}}{\emptysu}{\stuck}{\su}$.
\end{cor}

\begin{figure}[t]
\centering
\begin{mcode}
genWitness :: Nat -> [($x$, $e$)] -> [$v$]
genWitness n bs
       = [witness bs ($\vsu$ vs) | $\vsu$ <- $\vsu$s]
  where
   vs  = genArgs defs []      -- (1)
   $e$    = close defs vs        -- (2)
   res = take n (eval e)      -- (3)
   $\vsu$s   = [$\vsu$ | $\pair{\stuck}{\vsu}$ <- res]  -- (4)
\end{mcode}
\caption{Generating Witnesses}
\label{fig:algo-gen-witness}
\end{figure}

>>>>

\subsection{Syntax}
\label{sec:syntax}
\input{syntax.tex}
%
Figure~\ref{fig:syntax} describes the syntax of \lang, a simple lambda
calculus with integers, booleans, and binary trees.
%
As we are specifically interested in programs that \emph{do} go wrong,
we include an explicit \stuck\ configuration in our syntax.

\paragraph{Holes}
\label{sec:holes}
%
Recall that a key challenge in our setting is to find witnesses
that are meaningful and do not arise from choosing values from
irrelevant types.
%
We solve this problem be equipping our term language with a notion
of a \emph{hole}, written \ehole, which represents \emph{unconstrained}
values that may replaced with values from \emph{any} type.
%
Dually, holes may also appear in types, written \thole, where they may
be thought of as type variables that we will not generalize over.
%

@@

\subsection{Semantics}
\label{sec:semantics}
%

@@

\paragraph{Narrowing Types} The procedure $\force{v}{t}{\vsu}{\tsu}$,
defined in Figure~\ref{fig:narrow}, takes as input a value $v$ and
refines it to have the type $t$ by yielding a pair of either the same
value and value substitution, or yields the stuck term if no such
refinement is possible. In the case where $v$ is a hole, it first checks
in the given $\vsu$ to see if the hole has already been instantiated
and, if so, returns the existing instantiation.
%
\begin{figure*}[t]
$$
\begin{array}{lcl}
%% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
\forcesym                  & ::     & (v, t, \vsu, \tsu) \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu} \\
\force{v}{\thole}{\vsu}{\tsu}  & \defeq & \hspace{-1ex}
\begin{cases}
  \force{v}{\subst{\tsu}{\thole}}{\vsu}{\tsu} & \mbox{if}\ \thole \in dom(\tsu) \\
  \triple{v}{\vsu}{\extendsu{\tsu}{\thole}{\typeof{v}}} & \mbox{otherwise} \\
\end{cases} \\
\force{\vhole{t_1}}{t_2}{\vsu}{\tsu} & \defeq & \hspace{-1ex}
\begin{cases}
  \triple{v}{\vsu}{\tsu''}    & \mbox{if } v = \lookupsu{\vsu}{\ehole},
                                         \tsu' = \unify{t_1}{t_2}{\tsu},
                                         \tsu'' = \unify{\subst{\tsu'}{t_1}}
                                                        {\typeof{v}}{\tsu'} \\
  \triple{\stuck}{\vsu}{\tsu} & \mbox{if } v = \lookupsu{\vsu}{\ehole} \\
  \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu'} & \mbox{if}\ \tsu' = \unify{t_1}{t_2}{\tsu}, v = \gen{\subst{\tsu'}{t_1}}\\
\end{cases} \\
% \begin{cases}
%   \triple{\lookupsu{\vsu}{\ehole}}{\vsu}{\tsu} & \mbox{if}\ \ehole \in dom(\vsu), \hastype{\lookupsu{\vsu}{\ehole}}{t} \\
%   \triple{\stuck}{\vsu}{\tsu}                  & \mbox{if}\ \ehole \in dom(\vsu) \\
%   \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu}   & v = \gen{t}\\
% \end{cases} \\
 % \begin{array}{l}
    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \triple{\vsu(i)}{\vsu}\ \mathtt{else} \\
         % \elet{v}{\gen{t}}{\triple{v}{\ehole{i} \mapsto v}}
  % \end{array} \\
\force{n}{\tint}{\vsu}{\tsu}     & \defeq & \triple{n}{\vsu}{\tsu} \\
\force{b}{\tbool}{\vsu}{\tsu}    & \defeq & \triple{b}{\vsu}{\tsu} \\
\force{\efun{x}{e}}{\tfun}{\vsu}{\tsu} & \defeq & \triple{\efun{x}{e}}{\vsu}{\tsu} \\
\force{\vleaf{t_1}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vleaf{t_1}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{t_1}{t_2}{\tsu} \\
\force{\vnode{t_1}{v_1}{v_2}{v_3}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vnode{t_1}{v_1}{v_2}{v_3}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{t_1}{t_2}{\tsu} \\
\force{v}{t}{\vsu}{\tsu} & \defeq & \triple{\stuck}{\vsu}{\tsu}
\end{array}
$$
\caption{Narrowing values}
\label{fig:narrow}
\end{figure*}
%
While a hole may map to a value that \emph{contains} another hole, \eg a
lambda or a tree, it may not map \emph{directly} to another hole, so in
the second case above we do not need to \forcesym\ the result of the
substitution, the sub-hole will be narrowed when the context demands it.

\paragraph{Generating Values} The (non-deterministic) procedure $\gen{t}$
takes as input a type \texttt{t} and returns a value of that type. For base
types, the procedure returns an arbitrary value. For functions, it returns
a lambda with a \emph{new} hole denoting the return value, and unconstrained
types (denoted by $\thole$) yield fresh, unconstrained values
(denoted by $\ehole$).

$$
\begin{array}{lcll}
\gensym       & ::      & t \rightarrow v \\
\gen{\tint}   & \defeq  & n, & \quad \text{non-deterministic} \\
\gen{\tbool}  & \defeq  & b, & \quad \text{non-deterministic} \\
\gen{\ttree{t}}  & \defeq  & tr, & \quad \text{non-deterministic, well-typed} \\
\gen{\tfun}   & \defeq & \efun{x}{\ehole}, & \quad \text{\ehole is fresh} \\
\gen{\thole}  & \defeq & \vhole{\thole}, & \quad \text{\ehole is fresh} \\
\end{array}
$$

\paragraph{Steps and Traces}
\input{operational.tex}
%
% WRW notes that Figure 4 does not seem to handle recursion (it's not clear
% how the let rule would work for something "let rec"-y, and there's not
% function call rule). I only mention this because I can imagine a reviewer
% wondering about your ability to generate good witnesses for function
% types. This could likely be addressed in text, by a forward reference to
% Section 3.4 where higher-order functions are handled, without changing
% any of the formalisms at the last minute.
Figure~\ref{fig:operational} describes the small-step contextual
reduction semantics for \lang.
%
A \emph{closed expression} is a triple $\triple{e}{\vsu}{\tsu}$ of an expression $e$,
a value substitution $\vsu$, and a type substitution $\tsu$.
%
We write $\step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$ if the closed
expression $\triple{e}{\vsu}{\tsu}$ transitions in a \emph{small step} to
$\triple{e'}{\vsu'}{\tsu'}$.
%
A (finite) \emph{trace} $\trace$ is a sequence of closed expressions
$\triple{e_0}{\vsu_0}{\tsu_0}, \ldots, \triple{e_n}{\vsu_n}{\tsu_n}$ such that
$\forall 0 \leq i < n$, we have
$\step{e_i}{\vsu_i}{\tsu_i}{e_{i+1}}{\vsu_{i+1}}{\tsu_{i+1}}$.
%
We write \steptr{\trace}{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'} if $\trace$ is
a trace of the form $\triple{e}{\vsu}{\tsu},\ldots,\pair{e'}{\vsu'}{\tsu'}$.
%
We write \steps{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'} if
\steptr{\trace}{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'} for some trace $\trace$.

\paragraph{Primitive Reductions}
Primitive reduction steps --- addition, if-elimination, function
application, and data construction and case analysis --- use \forcesym
to ensure that values have the appropriate type (and that holes are
instantiated) before continuing the computation. Importantly,
beta-reduction \emph{does not} type-check its argument, it only ensures
that $v_1$ is a function.

@@

\paragraph{Witness Generality}
We now show that our evaluation relation instantiates holes in a
\emph{general} manner. That is, given a function $f$, if we have
$\steps{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,
then for every possible input type $t$, we can find a value
$\hastype{v}{t}$ such that $\eapp{f}{v}$ goes wrong.

\begin{thm}{\textbf{[Witness Generality]}}
\label{thm:soundness}
  For any function $f$, if\\
  \hbox{$\steptr{\trace}{\eapp{f}{\ehole}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,}
  then for every % inhabitable
  type
  % \footnote{We exclude builtin functions that subvert the type system, \eg \texttt{Obj.magic}, and thus consider the type $\forall a b. \tfun{a}{b}$ to be uninhabitable.}
  $t$ there exists $\hastype{v}{t}$ such that
  $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$.
\end{thm}

@@

\paragraph{The Type of an Expression} The (dynamic) type of an
expression $e$ is defined as a function $\typeof{e}$ defined as:
  \[
  \begin{array}{lcll}
    \typeof{n}   & \defeq & \tint & \\
    \typeof{b}   & \defeq & \tbool & \\
    \typeof{\efun{x}{e}} & \defeq & \tfun \\
    \typeof{\vleaf{t}} & \defeq & \ttree{t} \\
    \typeof{\vnode{t}{v_1}{v_2}{v_3}} & \defeq & \ttree{t} \\
    \typeof{\vhole{t}} & \defeq & t \\
    \typeof{e} & \defeq & \thole, & \quad \text{\thole is fresh} \\
  \end{array}
  \]
%
Note that we approximate the type of expressions rather than
performing full type inference, for example
$\typeof{\efun{x}{\eplus{x}{1}}} = \tfun$,
instead of the usual $\tint \rightarrow \tint$.

\paragraph{Dynamic Type Inference}
We can think of the evaluation of \eapp{f}{\ehole} as (dynamically)
inferring a \emph{partial type} --- a type that may contain holes ---
for $f$.
%
We can extract this type from evaluation traces.
%
Formally, we say that if
$\steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu}$,
then the \emph{partial type} of $f$ upto $\trace$, written \ptype{\trace}{f},
is $\typeof{\subst{\su}{\ehole{}}} \rightarrow \typeof{e}$.
%
%We will omit the subscript when we wish to refer to the final partial
%type, \ie\ at the step where the expression has been reduced to a value
%(or stuck.)

\paragraph{Type Compatibility}
Two types are compatible if one of them can be obtained by
suitably filling the holes of the other (\ie by suitably ``instantiating''
the type variables appearing in the other). Formally, a type $s$ is
\emph{compatible} with a type $t$, written \tcompat{s}{t}, if
$\exists \tsu.\ \subst{\tsu}{s} = \subst{\tsu}{t}$.

\paragraph{Preservation}
Given these two definitions, we show that each evaluation step
refines the partial type of $f$, \ie preserves type compatibility.
%
\begin{lem}
\label{lem:refine-partial}
If $\trace \defeq \triple{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu},\ldots$ and
$\trace' \defeq \trace, \triple{e'}{\vsu'}{\tsu'}$ (\ie $\trace'$ is a single-step
extension of $\trace$)
%
%The partial type of $f$ upto $\trace$ is compatible
%with the partial type upto $\trace'$, \ie\
%
then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
\end{lem}
\begin{proof}
  By induction on the length of $\trace$ and case analysis on the evaluation rules.
  %
  Note that all rules preserve partial types with the exception of when
  \forcesym\ is called on a hole, in which case we instantiate the hole with
  a concrete value.
  %
  But $\typeof{\ehole{}} = \thole{}$, which is compatible with any type.
\end{proof}

\paragraph{Narrowing}
%
Furthermore, only a call to \forcesym can change the partial type of $f$.
%
\begin{lem}
\label{lem:force-inst}
If
$\trace \defeq \triple{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$
and
$\trace' \defeq \trace, \step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$
and
$\ptype{\trace}{f} \neq \ptype{\trace'}{f}$,
then the final step $\step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$ invokes \forcesym.
\end{lem}

@@

\paragraph{Incompatible Values Are Wrong}
%
\emph{Any} value that is \emph{incompatible} with
the partial type upto trace $\trace$ will cause $f$ to get stuck
in \emph{at most} $k$ steps where $k$ is the length of $\trace$.
%
\begin{lem}
\label{lem:k-stuck}
  If \steptr{\trace}{\eapp{f}{\ehole}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu} and
     \hastype{v}{t} and
     \tincompat{t}{\ptype{\trace}{f}},
  then
     \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}.
\end{lem}
\begin{proof}
  By induction on $k$, the length of $\trace$.
  Let $s_{\trace} \rightarrow \_ = \ptype{\trace}{f}$ and
  $\trace' = \triple{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu},...,\triple{e'}{\vsu'}{\tsu'}$
  such that $\step{e'}{\vsu'}{\tsu'}{e}{\vsu}{\tsu}$.
  Suppose \hastype{v}{\tincompat{t}{s_{\trace}}}, we
  will show that \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}
  in at most $k$ steps:
%
  %\begin{description}
  % \item
  [Case \tincompat{t}{s_{\trace'}}:]
    The inductive hypothesis applies.
  %\item
  [Case $\tcompat{t}{s_{\trace'}}$ but $\tincompat{t}{s_{\trace}}$:]
    By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
    $k$.
    %
    Lemma~\ref{lem:refine-partial} tells us
    $\tcompat{s_{\trace'}}{s_{\trace}}$, which means we must have
    specifically narrowed $s_{\trace'}$ to a type incompatible with $t$.
    %
    A case analysis of the rules shows that \forcesym cannot have
    succeeded.
  %\end{description}
\end{proof}

@@

\paragraph{Modeling Semantics}
%: \texttt{eval} and \texttt{isStuck}}
%
We resolve the non-determinism in the operational semantics
(\S~\ref{sec:semantics}) via the procedure:
%
\ES{HERE}
\begin{mcode}
  eval :: ($e$, $\vsu$, $\tsu$) -> [($v$, $\vsu$, $\tsu$)]
\end{mcode}
%
Due to the non-determinism introduced by \gensym, a call
\hbox{|eval($e$, $\vsu$, $\tsu$)|} returns the \emph{list} of possible results
of the form $\triple{e'}{\vsu'}{\tsu'}$ such that
$\steps{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$.
%
We determine whether a given value is stuck by using the predicate:
%
\begin{mcode}
  isStuck :: $v$ -> Bool
\end{mcode}

@@

close :: [($x$, $e$)] -> [$v$] -> ($e$, $\vsu$, $\tsu$)
close bs vs = (e, $\emptysu$, $\emptysu$)
  where
    e       = mkLets bs (witness bs vs)

@@

\begin{cor}{\textbf{[Witness Generation]}}
\label{thm:generation}
  If @genWitness k [(f1,e1),...,(fn,en)]@ is non-empty,
  then for every type $t$ there exists $\hastype{v}{t}$ such that
  $\steps{\eapp{f_n}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$.
\end{cor}

\begin{figure}[t]
\centering
\begin{mcode}
genWitness :: Nat -> [($x$, $e$)] -> [$v$]
genWitness n bs
       = [witness bs ($\vsu$ vs) | $\vsu$ <- $\vsu$s]
  where
   vs  = genArgs defs []        -- (1)
   cfg = close defs vs          -- (2)
   res = take n (eval cfg)      -- (3)
   $\vsu$s   = [$\vsu$ | $\triple{\stuck}{\vsu}{\tsu}$ <- res]  -- (4)
\end{mcode}
\caption{Generating Witnesses}
\label{fig:algo-gen-witness}
\end{figure}
